---
title: "week 4 data analysis"
author: "mel"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(reshape2)
```

## Overview
This report presents preliminary data analysis for the week 3 example data. Specfically, this will report basic descriptive statisitcs as well as report tests of the following predictions.

1. People will feel more guilt when: 
  + 1.1 They caused the shock (vs. their partner)
  + 1.2 Their partne was morall good (vs. bad)
  +1.3 The shock was stronger (high > medium > low)
2. Wome will report more guilt tha men 
3. Agreeable people will report more guilt 
4. Nuerotic people will report more guilt 

and 

5. People will be more generous when: 
  + 5.1 They caused the shock (vs. their partner)
  + 5.2 The partner was morally good (vs. bad) 
  + 5.3 The shock was stronger (high > med > low)
6. Womne will be more generous than men 
7. Agrreable peopl will be more genrous 
8. Nuerotic people will be more generous 


```{r load data}
data <- read.csv("week3/processed data/Week 3 Example Data PROCESSED.csv")

#recode varibles 
#grab PIN vecotr and overide it with a categorlcal version of itself
data$PIN <-as.factor(data$PIN)
data$sex <-as.factor (data$sex)
data$shock <-as.factor(data$shock)
data$shockCause <- as.factor(data$shockCause)
data$pMoral <- as.factor( data$pMoral)
```

#descriptive statisitcs  for guilt: 
summary function: min, max, mean, quartiles
sd: standard deviation
na.rm : remove missing values 
```{r descGuilt}
c(summary(data$guilt), "SD" = sd(data$guilt,na.rm =T))

```
 # genersority descriptives
```{r descgenerosity}
#generosity before the shock
c(summary(data$preShare), "SD" = sd(data$preShare,na.rm =T))
#generosity after shock
c(summary(data$postShare), "SD" = sd(data$postShare,na.rm =T))

```
 
```{r descpersonality}
#agreeableness
c(summary(data$agreea), "SD" = sd(data$agreea,na.rm =T))
#Nuero
c(summary(data$nuero), "SD" = sd(data$nuero,na.rm =T))
```


#guilt analysis

### Prediction 1
```{r p1.1}
#run t-tes.for prediction 1.1
p1.1 <- t.test(data$guilt~data$shockCause)

#outout the result 
p1.1
```
Prediction 1.1 was supported
People who caused the shock felt more guilty when they caused the shock than when their partner did: 


```{r p1.2}
#Run the t.test fo prediction 1.2
p1.2 <- t.test(data$guilt~data$pMoral)
```
Prediction 1.2 was supported
People felt more guily when their partner was morally good

```{r p1.3}
#lm: linear model
#data = dataframe
# run the one-way anova for prediction for 1.3
p1.3 <- lm(guilt~shock, data=data)

#output the results 
anova(p1.3)

#output group means 
#help
tapply(data$guilt, data$shock,function(x) mean(x,na.rm=T))
```
prediction 1.3 was supported. people felt more guilty when their partner recived a stronger shock: 


#Prediction 2
t test
```{r p2}
#run the t-test for prediction 2
p2 <- t.test(data$guilt~data$sex)

#output the result 
p2
```
no sig difference
Prediction 2 was not supported. Women did not report more guilt than men

#Prediction 3
correlation
```{r p3}
#run the correlation test for prediction 3
p3 <- cor.test(data$guilt, data$agreea)

#output the result
p3
```
Prediction 3 was supported.  People who are more agreeable were more guilty
close to 1 stronger, closer to 0 weaker 

# Predcition 4
correlation
```{r p4}
# run the correlation test for prediction 3 
p4 <- cor.test(data$guilt, data$nuero)

#output the result 
p4
```
 Prediction 4 was not supported . Nuerotic people were not more guilty 
 
#Generosity Analysis

```{r melt}
#melt from wide to long

dataMelt <- melt(data,
                 measure.vars = c("preShare", "postShare"),
                 variable.name = c("prePost"),
                 value.name = c("share")
                 )


```

## Prediction 5.1
two by two anova
```{r p5.1}
#help
# run the within subject Anova for p5.1
p5.1 <-aov(share~prePost*shockCause+Error(PIN), data = dataMelt)

#output the results 
summary (p5.1)
 #output the means 
tapply(dataMelt$share,dataMelt[,c(24,21)], function(x) mean(x,na.rm=T)
       )
```
Prediction 5.1 was supported. people were more generous when they caused the shock and *less* generous when the partner caused the shock


